The investment strategy based on behavior of artificial earthworm for use in algotrading

### 2014-05-12 Article/Article20140406Granada/LatexArt/article_main.tex
layout fixes

\begin{abstract}
The paper presents an innovative concept of~investment strategy derived from the~general ideas of~artificial intelligence. The strategy has been tested in a~number of~simulations on historical data as well as data outside learning periods. Tests were performed mainly in the~selected currency markets, including the~primary currency pair of~EURUSD. Presented approach opens long and~short positions alternately during candle closing events. Closing a~position is based on recommendation which is deduced from changes in characteristics of~time series. The strategy has parameters that vary through time and~are adapted to a~market changes. To present the~strategy a~metaphor of~an innovative artificial earthworm that feeds on a~time series was used. The earthworm has its own artificial intelligence which enables it to analyse ,,consumed'' candles, change its preferences about movement direction or~to stay in neutral state. The goal was to implement presented strategy for trading platform to achieve the~automatic trading effect.  The strategy was tested in two development environments --- MATLAB environment and~trading platform MetaTrader after conversion of~M-file into MQL4 file. Implemented strategy enables to achieve interesting results.
\keywords{investment strategy, machine learning, forecasting, time series, algotrading, mql4}
\end{abstract}

@@

\end{itemize}
If this condition is fulfilled, then depending on the~current position status $PositionState$ (where $PositionState$ is a~variable which takes values from \{$long, short, none$\}) at the~opening of~the~next candle, a~position is opened or~continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =long \text{ then }PositionState =long;
\end{equation}
(costless continuation of~opened position);
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState = short \text{ then } PositionState=long;
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =none \text{ then } PositionState =long;
\end{equation}

@@

In formulas above there is a~parameter $p2$. It is the~second parameter of~the~strategy which is an arbitrarily separated part of~the~earthworm, which depending on size causes the~direction of~investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of~$p2$ leads to a~higher frequency of~these changes. The optimum value of~$p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a~single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the~$spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or~opening a~short position in case when there was no open position;\\
$cost$ is a~variable that depends on the~particular brokerage platform that was used for testing the~strategy; in this case, for the~currency pair EURUSD $cost = 0.00016$ was assumed and~for the~currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the~$spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or~opening a~long position in case when there was no open position.
\end{itemize}

@@

\begin{itemize}
\item To determine the~current quality of~the~strategy and~risk management summation of~current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the~current } i \text{ value;}
\end{equation}
\item There is a~possible local subsidence $Zc_i$ capital by~introducing two additional parameters to manage the~risk.\\
$p3$ --- the~number of~steps back on the~curve of~the~cumulative profit for the~determination of~$Zc_i$ --- $p3$ for comparison with the~present value of~$Zc_i$;
$p4$ --- maximum value measured in pips of~the~capital drawdown (eg for the~currency pair EURUSD), which if exceeded will trigger the~closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and~} PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where $PositionState$ --- is the~type of~open position (long or~short) and~$PositionState$ equal 0 means resignations from the~opening position to the~beginning of~the~next candle.
\end{itemize}
So, the~above-described algorithm uses 5 parameters and~3 of~them are devoted to risk management of~the~strategy.\\
%%PBStart
The complexity of~the~algorithm is illustrated in Figure \ref{fig:fig7}, which shows the~individual steps and~decisions made in the~iteration on a~single candle in the~form of~a~decision tree. This iteration was carried out both in the~search for optimal parameters $p1, ..., p5$ and~in the~process of~testing the~parameters on data that was not used in the~learning period.
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the~states and~decisions in one step (candle) in the~learning and~testing phase.}
\label{fig:fig7}
\end{figure}
\FloatBarrier

@@

To make sure that those market conditions are met in the~algorithm MATLAB script was converted manually to MQL4 expert advisor for trading platform of~MetaTrader 4 which is commonly used to perform the~real market operations.\\

@@

The concept of~an innovative investment strategy gave satisfactory results for a~certain pair as the~result of~periods of~machine learning. E.g. in the~studied data for two key market currency pairs sampled every hour, the~best value is 80 hours of~teaching and~15 hours of~testing. One can assume that this relationship may change in a~different environment (for other markets, for other periods) by~becoming a~sort of~next parameter that optimal values should be sought. It should be emphasized that the~study was conducted on data not used in the~learning phase and~for the~basic principle of~machine learning achieved results that would have satisfied many a~practice. It should also be noted that outside research conducted in environment completely specified (for constant unchanging data stored in MATLAB) was also performed simulations in terms of~actual brokerage platform. The results obtained in both environments are similar in the~sense of~direction changes and~logic behavior caused by~the~user received by~the~algorithm implemented in the~MATLAB and~MetaTrader. Differences were explained in this work.
The direction of~further research will test the~hypothesis that the~variability periods to test the~learner can improve the~quality of~prediction. This hypothesis seems to be interesting against the~background of~previous experiences and~results, eg for different studied strategies, including \cite{Wilinski2014}\cite{Wilinski}\cite{sinclare}\cite{krutsinger} obtained surprisingly different optimal length of~periods of~learning and~testing.

>>>>

\begin{abstract}
The paper presents an innovative concept of~investment strategy derived from the~general ideas of~artificial intelligence. The strategy has been tested in a~number of~simulations on historical data as well as data outside learning periods. Tests were performed mainly in the~selected currency markets, including the~primary currency pair of~\mbox{EURUSD}. Presented approach opens long and~short positions alternately during candle closing events. Closing a~position is based on recommendation which is deduced from changes in characteristics of~time series. The strategy has parameters that vary through time and~are adapted to a~market changes. To present the~strategy a~metaphor of~an innovative artificial earthworm that feeds on a~time series was used. The earthworm has its own artificial intelligence which enables it to analyse ,,consumed'' candles, change its preferences about movement direction or~to stay in neutral state. The goal was to implement presented strategy for trading platform to achieve the~automatic trading effect.  The strategy was tested in two development environments --- MATLAB environment and~trading platform MetaTrader after conversion of~M-file into MQL4 file. Implemented strategy enables to achieve interesting results.
\keywords{investment strategy, machine learning, forecasting, time series, algotrading, mql4}
\end{abstract}

@@

\end{itemize}
If this condition is fulfilled, then depending on the~current position status \textit{PositionState} (where $PositionState$ is a~variable which takes values from \{$long$, $short$, $none$\}) at the~opening of~the~next candle, a~position is opened or~continued as follows: 
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =long \text{ then }PositionState =long;
\end{equation}
(costless continuation of~opened position);
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState = short \text{ then } PositionState=long;
\end{equation}
\begin{equation}
\text{If } Rg=1 \text{ and~} PositionState =none \text{ then } PositionState =long;
\end{equation}

@@

In formulas above there is a~parameter $p2$. It is the~second parameter of~the strategy which is an arbitrarily separated part of~the~earthworm, which depending on size causes the~direction of~investment not to change. Large $p2$ causes long positions to change into short positions (or vice-versa) rarely. Small value of~$p2$ leads to a~higher frequency of~these changes. The optimum value of~$p2$ is chosen similarly like $p1$.
\begin{itemize}
\item 	When opening long position, profit after one candle (this does not mean that position was closed after a~single candle) can be calculated as:
\begin{equation}
Z_i = C_{i+1} - C_i - spread;
\end{equation}
where the~$spread = 0$, when opened long position is continued; \\
and $spread = cost$, when long position was opened after closing or~opening a~short position in case when there was no open position;\\
$cost$ is a~variable that depends on the~particular brokerage platform that was used for testing the~strategy; in this case, for the~currency pair EURUSD $cost = 0.00016$ was assumed and~for the~currency pair GBPUSD $cost = 0.00028$.
\item When short position is opened, profit after one candle (it does not mean that position is closed) can be calculated as: 
\begin{equation}
Z_i = C_i-C_{i+1}- spread;
\end{equation}
where the~$spread = 0$, when opened short position is continued;\\ 
and $spread = cost$, when short position was opened after closing or~opening a~long position in case when there was no open position.
\end{itemize}

@@

\begin{itemize}
\item To determine the~current quality of~the~strategy and~risk management summation of~current gains/losses (the accumulation) is performed
\begin{equation}
Zc_i = sum ( Z_i ) \text{ from } i = 1 \text{ to the~current } i \text{ value;}
\end{equation}
\item There is a~possible local subsidence $Zc_i$ capital by~introducing two additional parameters to manage the~risk.\\
$p3$ --- the~number of~steps back on the~curve of~the~cumulative profit for the~determination of~$Zc_i$ --- $p3$ for comparison with the~present value of~$Zc_i$;
$p4$ --- maximum value measured in pips of~the~capital drawdown (e.g. for the~currency pair EURUSD), which if exceeded will trigger the~closure position:\\
\begin{equation}
\text{If } ( Zc_i - p3 - Z_i ) <- p4 \text{ and~} PositionState \neq 0 \text{ then } PositionState = 0
\end{equation} 
where $PositionState$ --- is the~type of~open position (long or~short) and $PositionState$ equal 0 means resignations from the~opening position to the~beginning of~the~next candle.
\end{itemize}
So, the~above-described algorithm uses 5 parameters and~3 of~them are devoted to risk management of~the~strategy.\\
%%PBStart
The complexity of~the~algorithm is illustrated in Figure \ref{fig:fig7}, which shows the~individual steps and~decisions made in the~iteration on a~single candle in the~form of~a~decision tree. This iteration was carried out both in the~search for optimal parameters $p1, ..., p5$ and~in the~process of~testing the~parameters on data that was not used in the~learning period.
\begin{figure}[h!]
\centering
\includegraphics[width = \textwidth]{figures/rys7.png}
\caption{Decision tree for the~states and~decisions in one step (candle) in the~learning and~testing phase.}
\label{fig:fig7}
\end{figure}
\FloatBarrier

@@

To make sure that those market conditions are met in the algorithm \mbox{MATLAB} script was converted manually to MQL4 expert advisor for trading platform of~MetaTrader 4 which is commonly used to perform the~real market operations.\\

@@

The concept of~an innovative investment strategy gave satisfactory results for a~certain pair as the~result of~periods of~machine learning. E.g. in the~studied data for two key market currency pairs sampled every hour, the~best value is 80 hours of~teaching and~15 hours of~testing. One can assume that this relationship may change in a~different environment (for other markets, for other periods) by~becoming a~sort of~next parameter that optimal values should be sought. It should be emphasized that the~study was conducted on data not used in the~learning phase and~for the~basic principle of~machine learning achieved results that would have satisfied many a~practice. It should also be noted that outside research conducted in environment completely specified (for constant unchanging data stored in MATLAB) was also performed simulations in terms of~actual brokerage platform. The results obtained in both environments are similar in the~sense of~direction changes and~logic behavior caused by~the~user received by~the~algorithm implemented in the~MATLAB and~MetaTrader. Differences were explained in this work.
The direction of~further research will test the~hypothesis that the~variability periods to test the~learner can improve the~quality of~prediction. This hypothesis seems to be interesting against the~background of~previous experiences and~results, e.g. for different studied strategies, including \cite{Wilinski2014}\cite{Wilinski}\cite{sinclare}\cite{krutsinger} obtained surprisingly different optimal length of~periods of~learning and~testing.
